---
title: "Functions and Data Sources that Depend on SoilWeb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SoilWeb Functions and Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  # eval = !as.logical(Sys.getenv("R_SOILDB_SKIP_LONG_EXAMPLES", unset = "TRUE")),
  message = FALSE,
  warning = FALSE,
  fig.height = 7,
  fig.width = 7
)

# get source data for explanations
library(soilDB)
library(aqp)

x <- fetchOSD(c('miami', 'pierre', 'tristan', 'lucy', 'musick'), extended = TRUE)
```

# Summary
The soilDB package relies on a number of data sources curated as part of the [SoilWeb](https://casoilresource.lawr.ucdavis.edu/soilweb-apps) ecosystem of web applications, APIs, web coverage services (WCS), and internally-used lookup tables. These data are always updated shortly after the beginning of each fiscal year (October 1st) as part of the SSURGO refresh cycle. Within-cycle updates are typically performed quarterly. 

Many of these curated data sources are documented in:

  * O'Geen, A., Walkinshaw, M. and Beaudette, D.E. (2017), SoilWeb: A Multifaceted Interface to Soil Survey Information. Soil Sci. Soc. Am. J., 81: 853-862. https://doi.org/10.2136/sssaj2016.11.0386n
  
  * Beaudette, D.E. and O'Geen, A.T. (2010), An iPhone Application for On-Demand Access to Digital Soil Survey Information. Soil Sci. Soc. Am. J., 74: 1682-1684. https://doi.org/10.2136/sssaj2010.0144N
  
  * Beaudette, D.E. and O’Geen, A.T. (2009), Soil-Web: An online soil survey for California, Arizona, and Nevada, Comp. & Geo. Sci., 35:2119-2128, https://doi.org/10.1016/j.cageo.2008.10.016


## Primary Data Sources

  * [SSURGO](https://www.nrcs.usda.gov/resources/data-and-reports/soil-survey-geographic-database-ssurgo), the detailed soil survey of the United States
  * [STATSGO2](https://www.nrcs.usda.gov/resources/data-and-reports/description-of-statsgo2-database), the generalized soil survey of the United States
  * Official Series Descriptions (OSD) via [SoilKnowledgeBase](https://github.com/ncss-tech/SoilKnowledgeBase/)
  * Soil Classification database (no public version)
  * [USDA Ag. Handbook 296 -- Major Land Resource Area (MLRA)](https://www.nrcs.usda.gov/resources/data-and-reports/major-land-resource-area-mlra)
  * Climate ([PRISM](https://prism.oregonstate.edu/)) data and derivatives
  * Digital elevation model (DEM) derivatives
  * [NCSS Lab Data Mart](https://ncsslabdatamart.sc.egov.usda.gov/) snapshot
  * NRCS [Rapid Carbon Assessment](https://www.nrcs.usda.gov/resources/data-and-reports/rapid-carbon-assessment-raca) (RaCA)


## `soilDB` Functions

 * `fetchOSD()`: get soil morphology  and many soil series level summaries for one or more soil series names
 * `OSDquery()`: search the OSD records using the postgresql full text query syntax
 * `siblings()`: get names and basic information about soil series that co-occurr within soil map units containing  a given series
 
 * `seriesExtent()`: get vector or raster representations of were soil series have been used in SSURGO
 * `taxaExtent()`: get raster representations of were taxa and formative elements (Soil Taxonomy) have been used in SSURGO
 
 * `mukey.wcs()`: get raster data describing map unit keys, from a variety of sources, by bounding-box
 * `ISSR800.wcs()`: get raster data describing generalized patterns in soil property and condition, by bounding-box
 * `soilColor.wcs()`: get raster soil color maps by bounding-box
 
 

 
### Functions Near Retirement

 * `fetchKSSL()`: get soil characterization and (limited) morphology data
 * `fetchRaCA()`: get records associated with the RaCA collection
 * `SoilWeb_spatial_query()`: simple interface to spatial intersection between SSURGO polygons and user supplied bounding-box




## Related APIs

 * https://soilmap4-1.lawr.ucdavis.edu/osd-search/index.php
 * https://casoilresource.lawr.ucdavis.edu/osd-search/search-entire-osd.php
 




# SoilWeb Curated Data Sources


## SSURGO Derivatives


### MLRA

```{r}
knitr::kable(head(x$mlra), digits = 2)
```


### Geomorphic Summaries

```{r}
knitr::kable(head(x$hillpos), digits = 2)

knitr::kable(head(x$geomcomp), digits = 2)

knitr::kable(head(x$mtnpos), digits = 2)

knitr::kable(head(x$terrace), digits = 2)

knitr::kable(head(x$flats), digits = 2)

knitr::kable(head(x$shape_across), digits = 2)

knitr::kable(head(x$shape_down), digits = 2)
```

### Other

```{r}
knitr::kable(head(x$NCCPI), digits = 2)

knitr::kable(head(x$ecoclassid), digits = 2)
```


### Parent Material Summaries

```{r}
knitr::kable(head(x$pmkind), digits = 2)

knitr::kable(head(x$pmorigin), digits = 2)
```



## Climate Data and Derivatives

```{r}
knitr::kable(head(x$climate.annual), digits = 0)

knitr::kable(head(x$climate.monthly), digits = 0)
```



## DEM Derivatives

```{r}
knitr::kable(head(x$geomorphons), digits = 2)
```


## SC and OSD Derivatives

```{r}
knitr::kable(head(x$competing))

knitr::kable(head(x$geog_assoc_soils))
```



Relevant SoilWeb data metadata, as reported by `fetchOSD(..., extended = TRUE)`:


|product                       |last_update |
|:-----------------------------|:-----------|
|ISSR800                       |2024-10-30  |
|KSSL snapshot                 |2025-01-21  |
|MLRA membership               |2025-10-07  |
|OSD fulltext                  |2026-01-02  |
|OSD morphology                |2026-01-02  |
|SC database                   |2026-01-02  |
|series climate summary        |2025-10-08  |
|series-ESID cross-tabulation  |2025-10-07  |
|SSURGO geomorphology          |2025-10-06  |
|SSURGO geomorphon             |2026-01-11  |
|SSURGO NCCPI Stats            |2025-10-19  |
|SSURGO parent material        |2025-10-06  |
|series extent                 |2025-10-04  |
|taxa extent                   |2025-10-21  |




# Relevant Tutorials




# Examples
```{r eval=FALSE}
library(soilDB)
library(aqp)

x <- fetchOSD(c('miami', 'pierre', 'tristan', 'lucy', 'musick'), extended = TRUE)


```




